Augmentation of a Historical Harpsichord Keyboard Replica for Haptic-Enabled Interaction in Museum Exhibitions

### 2025-01-24 src/main.tex

%%
%% This is file `sample-nime-paper.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,sigconf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-nime-paper.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% History of NIME Template Updaters:
%% Modified by Florent Berthaut, Charles Martin, Yichen Wang, 21 November 2024
%% Modified by Adnan Marquez-Borbon 30 November 2022
%% Modified by Courtney Reed 28 November 2022
%% Modified by Joe Wright 14 December 2019
%% Modified by Niccolò Granieri 10 October 2018
%% Modified by Angelo Fraietta 23 December 2018
%% Modified by Angelo Fraietta 22 November 2018
%% Modified by Rodrigo Schramm on 22 September 2018
%% Modified by Luke Dahl on 17 October 2-17
%% Modified by Cumhur Erkut on <2016-10-11 Tue>
%% Modified by Edgar Berdahl on 5 November 2014
%% Modified by Baptiste Caramiaux on 25 November 2013
%% Modified by Kyogu Lee on 7 October 2012
%% Modified by Georg Essl on 7 November 2011
%%
%% And here's the template:
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{nimeart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{nimeart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[sigconf]{nimeart}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}

@@

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Matthew Hamilton}
% \authornote{Both authors contributed equally to this research.}
\email{matthew.hamilton2@unibo.it}
% \orcid{1234-5678-9012}
\author{Michele Ducceschi}
% \authornotemark[1]
\email{michele.ducceschi@unibo.it}
\affiliation{%
  \institution{Università di Bologna}
  \city{Bologna}  
  \country{Italy}
}

\author{Roberto Livi}
\author{Catalina Vincens}
\affiliation{%
  \institution{Museo di San Colombano}
  \city{Bologna}
  \country{Italy}}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
This work describes the design of an electronically augmented replica of a
16th-century harpsichord keyboard with a typical Renaissance Italian layout. The
keyboard is a collaboration between the \anon{NEMUS project} and the Museo di San Colombano,
which houses the Tagliavini Collection of playable historical keyboard
instruments. The keyboard was commissioned for exhibition purposes and to
enhance the experiential authenticity of the collection. The keyboard emulates
the mechanics of a two-register harpsichord featuring a single manual. Unpitched
strings under tension replicate the mechanical resistance characteristic of the
original compass. An array of infrared LED / phototransistor pairs, connected to a
microcontroller, was installed within the interface to generate a continuous
stream of data corresponding to the jacks' motion. This paper details the design
principles, development, open source implementation, and execution of the installation for the San Colombano exhibition.
\end{abstract}

@@

% Option 2:
\begin{teaserfigure}
 \hspace*{0.6in}
  \includegraphics[width=0.8\textwidth]{src/images/49-key-front-top.png}
  \caption{49-key Model Harpsichord Mechanism by \anon{Roberto Livi, Bologna 2024}}
  \Description{49-key Model Harpsichord Mechanism by \anon{Roberto Livi, Bologna 2024}}
  \label{fig:teaser}
\end{teaserfigure}

@@

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
A project of this breadth developed across 3 institutes and 2 countries
cannot be carried out without the support and technical knowledge of many
skilled and passionate people.

We acknowledge the invaluable contributions of

\begin{itemize}
\item
  Università di Bologna
  \begin{itemize}
  \item
    Alma Labor
    \begin{itemize}
    \item
      Massimiliano Fraulini and the staff of Alma Labor at the Università
      di Bologna for their 3D printing expertise and generosity with
      electronics equipment
    \end{itemize}
  \item
    Fisica Tecnica
    \begin{itemize}
    \item
      Maurzio Chendi Fabrizio Casarini, Giulio Severo and Barbara
      Costantini
    \end{itemize}
  \end{itemize}
\item
  San Colombano
  \begin{itemize}
  \item
    Catalina Vincens
  \item
    Roberto Livi
  \end{itemize}
\item
  Edinburgh
  \begin{itemize}
  \item
    uCreate
    \begin{itemize}    
    \item
      Simeon, Stuart, Millie, and Ruth of uCreate at the Univeristy of
      Edinburgh for their support with soldering facilities.
    \end{itemize}
  \item
    University of Edinburgh
    \begin{itemize}
    \item
      Richard Collins and the Digital Making Staff at the Edinburgh
      College of Art for 3D printing instruction and CNC support.
    \item
      Joe Hathaway of Digital Development at the College of Art for
      providing electronics components
    \item
      St Cecilia's Jenny Nex
    \end{itemize}
  \end{itemize}
\end{itemize}

>>>>

%%
%% This is file `sample-nime-paper.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,sigconf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-nime-paper.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% History of NIME Template Updaters:
%% Modified by Florent Berthaut, Charles Martin, Yichen Wang, 21 November 2024
%% Modified by Adnan Marquez-Borbon 30 November 2022
%% Modified by Courtney Reed 28 November 2022
%% Modified by Joe Wright 14 December 2019
%% Modified by Niccolò Granieri 10 October 2018
%% Modified by Angelo Fraietta 23 December 2018
%% Modified by Angelo Fraietta 22 November 2018
%% Modified by Rodrigo Schramm on 22 September 2018
%% Modified by Luke Dahl on 17 October 2-17
%% Modified by Cumhur Erkut on <2016-10-11 Tue>
%% Modified by Edgar Berdahl on 5 November 2014
%% Modified by Baptiste Caramiaux on 25 November 2013
%% Modified by Kyogu Lee on 7 October 2012
%% Modified by Georg Essl on 7 November 2011
%%
%% And here's the template:
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{nimeart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{nimeart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[sigconf]{nimeart}
\usepackage{graphicx,subcaption}
% \usepackage{tikz}
% \usetikzlibrary{shapes,arrows}

@@

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Matthew Hamilton}
% \authornote{Both authors contributed equally to this research.}
\email{matthew.hamilton2@unibo.it}
% \orcid{1234-5678-9012}
\author{Michele Ducceschi}
% \authornotemark[1]
\affiliation{%
  \institution{Università di Bologna}
  \city{Bologna}  
  \country{Italy}
}

\author{Roberto Livi}
\affiliation{%
  \institution{Harpsichord builder}
  \city{Pesaro}
  \country{Italy}
  }

\author{Catalina Vincens}
\affiliation{%
  \institution{Museo di San Colombano}
  \city{Bologna}
  \country{Italy}}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
This work describes the design of an electronically augmented replica of a
16th-century harpsichord keyboard with a typical Renaissance Italian layout to create a DMI. The
project is a collaboration between the \anon{NEMUS project} and the Museo di San Colombano,
which houses the Tagliavini Collection of playable historical keyboard instruments. 
The keyboard was commissioned for exhibition where it will act as an interface to  digitised instruments in the collection.

The project aims to enhance the experiential authenticity of the collection and address how we can conserve the interaction with historical instruments when the interface is a limiting factor in the verisimilitude of the experience.

The tactile feedback of the keyboard is created by incorporating the same mechanics of a two-register harpsichord with a single manual. 
Unpitched strings under tension replicate the mechanical resistance characteristic of the original compass. 
An array of optical sensors connected to a microcontroller was installed within the interface to generate continuous
streams of data corresponding to the vertical motion of each jack. 
This paper details the keyboard's design principles, technical development, and the motivations behind them. 
% It also discusses the implications of its execution for installation for public exhibition. 
\end{abstract}

@@

% Option 2:
\begin{teaserfigure}
 \centering
  \includegraphics[width=0.58\textwidth]{src/images/49-key-front-top.png}
  \caption{49-key Model Harpsichord Mechanism by \anon{Roberto Livi, Bologna 2024}}
  \Description{49-key Model Harpsichord Mechanism by \anon{Roberto Livi, Bologna 2024}}
  \label{fig:teaser}
\end{teaserfigure}

@@

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
This work received funding from the European Commission's Horizon2020 programme, through grant ERC-StG 950084 NEMUS. We acknowledge the invaluable contributions of
\begin{itemize}
\item
  Università di Bologna
  \begin{itemize}
  \item
    Alma Labor
    \begin{itemize}
    \item
      Massimiliano Fraulini and the staff of Alma Labor at the Università
      di Bologna for their 3D printing expertise and generosity with
      electronics equipment
    \end{itemize}
  \item
    Fisica Tecnica
    \begin{itemize}
    \item
      Maurzio Chendi Fabrizio Casarini, Giulio Severo and Barbara
      Costantini
    \end{itemize}
  \end{itemize}
\item
  Edinburgh
  \begin{itemize}
  \item
    uCreate
    \begin{itemize}    
    \item
      Simeon, Stuart, Millie, and Ruth of uCreate at the Univeristy of
      Edinburgh for their support with soldering facilities.
    \end{itemize}
  \item
    University of Edinburgh
    \begin{itemize}
    \item
      Richard Collins and the Digital Making Staff at the Edinburgh
      College of Art for 3D printing instruction and CNC support.
    \item
      Joe Hathaway of Digital Development at the College of Art for
      providing electronics components
    \item
      St Cecilia's Jenny Nex
    \end{itemize}
  \end{itemize}
\end{itemize}
